---
output:
  rmarkdown::html_vignette: default
  html_document: default
---

The `DABOM` compendium can be downloaded as a zip from from this URL: <https://github.com/BiomarkABS/DABOM/archive/master.zip>

Or the user can install the compendium as an R package from GitHub by using Hadley Wickham's `devtools` package:

```{r gh-installation, eval = FALSE}
# install and load remotes, if necessary
install.packages("devtools")
devtools::install_github("KevinSee/DABOM", 
                         build_vignettes = TRUE)
```

`devtools` may require the downloading and installation of Rtools. The latest version of Rtools can be found [here](https://cran.r-project.org/bin/windows/Rtools/).

For the latest development version:

```{r develop-version, eval = FALSE}
devtools::install_github("KevinSee/DABOM@develop")
```

### JAGS Software

The user will also need the [JAGS](http://mcmc-jags.sourceforge.net/) software to run DABOM. They can download that from [SourceForge](https://sourceforge.net/projects/mcmc-jags/files/). JAGS (**J**ust **A**nother **G**ibbs **S**ampler) software is used by `DABOM` for Bayesian inference. Please download version >= 4.0.0.

### Other R Packages

The user will also need to install `tidyverse`, a series of R packages that work together for data science (i.e. data cleaning and manipulation), as well as the `rjags` package to interface with JAGS. The `tidyverse` and `rjags` packages are all available from the R community and can be installed by typing the following into your R console:

```{r install-cran, eval = F}
install.packages("tidyverse")
install.packages("rjags")
```
